---
title: "The Environmental Impact of Oil Pipeline Spills"
author: "Caitlin Timmons, Sarah Weden, Kendall Al-Bashir"
date: "Last updated on `r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    theme: paper
    toc: true
    toc_depth: 2
    toc_float: true
    df_print: kable
---

```{r, echo=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, message = FALSE, warning = FALSE, 
  fig.width = 16/2, fig.height = 9/2)
```

<!--
Comments in HTML are like this! 
-->




# "Liquid Gold": Oil's Expense On the Environment

The following scatterplot demonstrates how, for a representative sample of pipeline spills, the amount of oil released in the spill directly correlates to the costs paid out by the responsible corporation to repair environmental damage. This reinforces what is already known about the severe damage that oil poses to ecological systems, and shows that the amount of oil released can be used as another proxy by which to quantify the environmental hazard of any particular pipeline spill. 


``` {r, echo =TRUE, message =FALSE, warning =FALSE, fig.width = 10, fig.height = 10}
library(tidyverse)
library(leaflet)
library(sf)
library(readxl)
library(maps)
library(USAboundaries)
library(scales)

us_counties <- us_counties(map_date = NULL, resolution = c("high"), states = NULL) %>%
  select(name, state_abbr) %>%
  mutate("State_ID" = state_abbr, "County" = name) %>%
  filter(State_ID != "GU", State_ID != "VI", State_ID != "PR", State_ID != "AS", State_ID != "MP", State_ID != "HI", State_ID != "AK") 


USA_sf <- maps::map("state", plot = FALSE, fill = TRUE) %>%
  sf::st_as_sf() %>%
  sf::st_transform(4326) %>%
  mutate("State" = ID) %>%
  select(-ID)


pipeline_spills <- pipeline_spills <- read_excel("database.xlsx") %>%
  unite(col = "Accident_Location", c("Accident_City", "Accident_State"), sep = ", ", remove = FALSE) %>%
  mutate("State_ID" = Accident_State, "Barrels_Gallons" = Unintentional_Release_Barrels*42) %>%
  separate(col = Operator_Name, into = c("Operator_General"), sep = " ", remove = FALSE, convert = FALSE, 
           extra = "drop", fill = "warn") %>%
  filter(Environmental_Remediation_Costs > 0, Longitude < -60)

```



```{r, echo =TRUE, message =FALSE, warning =FALSE, fig.width = 10, fig.height = 8}

pipeline_operators <- pipeline_spills %>%
  group_by(Operator_General) %>%
  summarize(n_spills = n()) %>%
  filter(n_spills > 5) %>%
  inner_join(pipeline_spills, by = "Operator_General") 

ggplot(data = pipeline_spills, mapping = aes(x = Barrels_Gallons, y = Environmental_Remediation_Costs)) +
  geom_jitter(width = 0.3, height = 0.3, alpha = 0.6) +
  geom_smooth(method = lm) +
  scale_y_log10(label=comma) +
  scale_x_log10(label=comma) + 
  labs(title = "Correlation between amount of oil spillage and cost to the environment", subtitle = "Gallons of oil released in each pipeline spill versus the environmental remediation costs of the spill. 
Regression line demonstrates positive trend.", x = "Oil Released (Gallons)", y = "Environmental Remediation Costs ($)")

```

# Who Is Responsible for the Environmental Damage?

We already explored in our initial map some of the corporations which are causing pipeline spills. But not all of these companies are contributing equally to environmental damage. The following barplot reveals the corporations most culpable for environmental damage, by calculating the total amount of oil they have released into the environment. All of these companies are “repeat offenders,” given that they caused more than 5 spills in this timeframe. Be sure to note the scale on the x-axis, which represents the total gallons of oil released into the environment. The numbers are quite large, meaning even corporations with the smallest bars are still causing significant pollution. 


```{r, echo =TRUE, message =FALSE, warning =FALSE, fig.width = 10, fig.height = 8}

pipeline_barrels <- pipeline_operators %>%
  group_by(Operator_General) %>%
  summarize(Sum_Gallons = sum(Barrels_Gallons)) %>%
  mutate(Operator_General = reorder(Operator_General, -Sum_Gallons)) %>%
  filter(Sum_Gallons > 1000)

ggplot(data = pipeline_barrels, mapping = aes(x = Operator_General, y = Sum_Gallons)) +
  geom_col(fill = "red") +
  coord_flip() +
  scale_fill_discrete(guide = FALSE) +
  labs(title = "Which Corporations Spilled the Most Oil?", subtitle = "Barplot representing the total gallons of oil released by most culpable corporations from 2010-present.", y = "Total Oil Released (Gallons)", x = "Corporation Name")

```


# Pipeline Spills and Overall Pollution


We then wanted to examine how these pipeline spills fit in with the larger scheme of pollution in different areas of the country. This information is relevant given the far-reaching and long-lasting effects of pipeline spills; they may be contributing to or worsening existing pollution. **The following map uses data collected from the U.S Environmental Protection Agency to map the total ozone release levels (in blue) by county between 2000 and 2016.** Note that the unit “parts per billion” is a measure of volume which denotes the concentration of pollutants in the air. Pipeline spills are marked by pink dots.

Although not representative of every county, we can see that especially on the West coast, areas with significant oil spills overlap with high levels of pollution, meaning that these pipelines are damaging an already damaged system. Moreover, in underdeveloped areas of the country with higher ozone release levels, it is likely that these pipelines are contributing significantly to that pollution. 



``` {r, echo =TRUE, message =FALSE, warning =FALSE, fig.width = 14, fig.height = 14}

pollution <- read_excel("pollution.xlsx") %>%
  separate(col = Date_Local, into = c("Year", "Month", "Day"), sep = "-", 
           convert = FALSE, extra = "warn", fill = "warn") %>%
  group_by(County) %>%
  summarize(Sum_O3 = sum(O3_Mean)) %>%
  right_join(us_counties, by = "County") %>%
  select(-c(name, state_abbr)) %>%
  drop_na()

pipeline_spills_sf <- pipeline_spills %>%
  filter(Longitude >= -130) %>%
  st_as_sf(coords = c("Longitude", "Latitude"), crs = 4326) 

ggplot() +
  geom_sf(data = USA_sf) +
  geom_sf(data = pollution, aes(fill = Sum_O3)) +
  scale_fill_continuous(low = "lightblue3", high = "navy", 
                        space = "Lab", na.value = "white",         
                        guide = "colourbar", aesthetics = "fill") +
  geom_sf(data = pipeline_spills_sf, size = 0.7, col = "deeppink2") +
  coord_sf(crs = 4326) +
  labs(title = "Relationship Between Pollution Levels and Pipeline Spills", 
       subtitle = "Total ozone release by county from 2000-2016, plotted against occurrances of oil pipeline spills from 2010-present", x = "Longitude", y = "Latitude", fill = "Total O3 Release (Parts Per Billion)")

```

# Conclusion

We’ve learned that despite a lack of thorough media coverage, oil spills in the U.S are very prevalent, with far-reaching environmental impacts. Corporations have been spilling hundreds of thousands of of gallons of oil into the environment, and despite paying out huge sums of money for environmental remediation, their cleanup efforts are often incomplete This leads to long-term ecological damage and increased overall pollution levels. Moreover, many of these corporations are “repeat offenders;” Enterprise alone caused 154 spills from 2010-2017. And with the U.S rising to the world’s largest oil producer, these spills will continue to happen. 

If you’re motivated to fight back against the corporations harming the environment, consider donating or volunteering to support organizations like the Water Is Life Movement, League of Conservation Voters, the Sierra Club, and the U.S Climate Action Network, among others.
